\documentclass[
  spinewidth=0.8325in,
  coverwidth=6in,
  coverheight=9in,
  marklength=0.125in,
  bleedwidth=0.125in,
  11pt
]{bookcover}

\usepackage{lipsum}

\newbookcovercomponenttype{center rotate}{
    \parbox[t][\partheight][c]{\partwidth}{
        \begin{center}
             \rotatebox[origin=c]{-90}{#1}
        \end{center}}}

\definecolor{paleblue}{RGB}{175,238,238}
\definecolor{darkblue}{RGB}{0,0,139}
\definecolor{darkerblue}{RGB}{0,0,40}

\begin{document}
\begin{bookcover}

\bookcovercomponent{color}{bg whole}{color=paleblue}

% Text on the front cover
\bookcovercomponent{normal}{front}{
    \centering
    \vspace{60mm}
    \color{black}
    {\sffamily\bfseries \resizebox*{50mm}{8mm}{Yael Medini}}
    \par\vspace{20mm}
    %\resizebox*{90mm}{40mm}{\parbox{35mm}{\centering Pale Blue Valley}}
    % {\bfseries\fontsize{40}{44}\selectfont  Pale Blue Valley}
    \resizebox*{100mm}{14mm}{\bfseries Pale Blue Valley}
}

% Text on the spine
\bookcovercomponent{center rotate}{spine}{%
    \color{black}\huge\sffamily\bfseries%
    Yael Medini \hspace{1in} \Huge Pale Blue Valley}

\bookcovercomponent{normal}{back}{
    \vspace{0.6in}
    \centering
    \parbox{.8\textwidth}{\color{darkerblue}
\normalsize

Pale Blue Valley is the story of a love-triangle in no specific place
or time, involving a woman and two men. The narrative unfolds against
a background of political unrest between a dominant ruling entity and
a resentful minority called Balad who once lived in a place known as
Pale Blue Valley.  This unrest finds expression in acts of terror
organized by the minority, which in turn provoke harsh military
repression.

Aera -- the young woman in the triangle -- is married to Yamik. The
couple are Balads, and Yamik is a member of his people's underground
Freedom Fighters. Glin, the\ other young man in the triangle, is
totally immersed in the world of music, belongs with the prevailing
dominant entity.

A secret love affair springs up between Aera and Glin.

Glin and Yamik happen to make each other's acquaintance, neither
having an inkling of Aera's connection with the other.

Aera is torn between the two. Her inner struggle is exacerbated when
the men's acquaintanceship turns into a warm friendship. Through his
friendship with Yamik Glin is drawn into and excited by the elevated
ideals of the Balads; they remind him of the wonders of music. He is
ready to join Yamik in his terrorist activities.

Aera gives birth to a baby whose paternity is unclear. Each of the men
believes the child to be his.

The unrelenting political conflict between the two entities culminates
in~a major terror event,\ causing over a hundred\ deaths. The three
protagonists of the ``triangle'' are among them. One of its only two
survivors is Aera's baby. No one comes to claim him nor wants to adopt
him\ since he is thought to be of Balad ethnicity. In time he is
adopted by an elderly couple~belonging to the ruling entity.

Aera's mother tracks down her baby grandson and comes to his adoptive
home. Her offer to help with the housework there and be the baby's
nanny is accepted.

\bigskip

She identifies herself as a Balad, but does not as yet reveal her
blood tie with the baby.

A period of political calm is expected. Who can tell for how long?
And who knows what the future holds for the baby?

    }


}


\end{bookcover}
\end{document}
